---
title: "Enrichment analysis"
output: 
  html_notebook:
    code_folding: hide
---


# TODO:
- create GSEA excel table
- do GO terms analysis


```{r,include=FALSE}
library(SummarizedExperiment)
library(magrittr)
library(ggplot2)
library(patchwork)
```


# Load data, and add manual geneset annotations
```{r}
x <- readRDS("data_with_go.rds")
rownames(x) <- make.unique(mcols(x)$gene)
mcols(x)$manual_genesets <- local({
  G <- read.csv2("manual_genesets.csv")
  G$locus_tag <- factor(G$locus_tag,unique(mcols(x)$locus_tag))
  unique(splitAsList(G$category,G$locus_tag))[mcols(x)$locus_tag]
})
mcols(x)$kegg_genesets <- local({
  G <- read.csv2("kegg_genesets.csv")
  G$locus_tag <- factor(G$locus_tag,unique(mcols(x)$locus_tag))
  unique(splitAsList(G$Category,G$locus_tag))[mcols(x)$locus_tag]
})
```


# Add sigE annotation to regDB
```{r}
is_sigE <- any(grepl("^Sigma E",mcols(x)$manual_genesets))
mcols(x)$regdb[is_sigE] <- pc(rep(CharacterList("SigE"),sum(is_sigE)),mcols(x)$regdb[is_sigE])
```



# Compute some summary statistics
```{r}
x <- local({
  # Log transform RPKM read counts matrix
  m <- log2(assay(x,"rpkm")+1)
  # Split the whole matrix, into submatrices of replicates
  m <- splitAsList(colnames(m),sprintf("%s%.3d",x$cond,x$time)) %>%
    lapply(function(cn) m[,cn])
  
  # Compute average replicate values
  mcols(x)$avg_log2_rpkm <- sapply(m,rowMeans)

  # List the pairwise comparison to perform, and compute the log2_fc
  cmp <- matrix(byrow=TRUE,ncol=2,c(
    "C005","C000",
    "C030","C000",
    "C120","C000",
    "C030","C005",
    "C120","C005",
    "C120","C030",
    "T005","C000",
    "T030","C000",
    "T120","C000",
    "T030","T005",
    "T120","T005",
    "T120","T030")
  )
  mcols(x)$log2_fc_avg <- mcols(x)$avg_log2_rpkm[,cmp[,1]] - mcols(x)$avg_log2_rpkm[,cmp[,2]]
  colnames(mcols(x)$log2_fc_avg) <- paste0(cmp[,1],"/",cmp[,2])
  x
})
```



# Output the excel table
```{r}
local({
  mcols(x)$count <- assay(x,"count_antisense")
  mcols(x)$rpkm <- assay(x,"rpkm")
  mcols(x)$position <- as.character(granges(x))
  D <- mcols(x)[c("position","locus_tag","gene","log2_fc_avg","avg_log2_rpkm","rpkm","count","manual_genesets","regdb")]
  D$manual_genesets <- unclass(table(D$manual_genesets))
  D$regdb <- unstrsplit(D$regdb,";")
  write.table(as(D,"data.frame"),sep="\t",file = "out/gene_count.tsv",row.names=FALSE,na = "")
})
```



# Make scatter plots comparing all genes expressions between 2 conditions
```{r}
ggscatter <- function(x,qx,qy) {
  mcols(x)[c("locus_tag","gene","log2_fc_avg")] %>%
    as("data.frame") %>%
    ggplot(aes_(x=qx,y=qy)) + 
      coord_equal(xlim=c(-10,10),ylim=c(-10,10)) +
      geom_point(aes(
        text = gene,
        color = abs(..x.. - ..y..) < log2(1.5)
      ),size=0.6) +
      geom_abline(slope=1,intercept = 0) +
      scale_color_manual(values=c("#2C7FB8","#7FCDBB")) +
      theme(legend.position="none",panel.grid = element_blank(),panel.background = element_blank())
}
p0 <- ggscatter(x,~`log2_fc_avg.C005/C000`,~`log2_fc_avg.T005/C000`) + ggtitle("0' >> 5'")
p1 <- ggscatter(x,~`log2_fc_avg.C030/C005`,~`log2_fc_avg.T030/T005`) + ggtitle("5' >> 30'")
p2 <- ggscatter(x,~`log2_fc_avg.C120/C030`,~`log2_fc_avg.T120/T030`) + ggtitle("30' >> 120'")
p0+p1+p2
ggsave("out/scatter.pdf",width=15,height=6)

plotly::ggplotly(p0,tooltip="text") %>% plotly::config(displayModeBar = FALSE)
plotly::ggplotly(p1,tooltip="text") %>% plotly::config(displayModeBar = FALSE)
plotly::ggplotly(p2,tooltip="text") %>% plotly::config(displayModeBar = FALSE)
```



```{r}
ggmap <- function(x,qx,qy,highlight) {
  highlight <- as.logical(highlight)
  X <- mcols(x)[c("locus_tag","gene","log2_fc_avg")] %>%
    cbind(highlight_mask=highlight_mask) %>%
    as("data.frame")
  ggplot(mapping = aes_(x=qx,y=qy)) + 
      coord_equal(xlim=c(-10,10),ylim=c(-10,10)) +
      geom_point(aes(text = gene),size=0.6,color="#7FCDBB",data=X[!X$highlight_mask,]) +
      geom_point(aes(text = gene),size=0.6,color="#2C7FB8",data=X[X$highlight_mask,]) +
      geom_abline(slope=1,intercept = c(-log2(1.5),log2(1.5))) +
      theme(legend.position="none",panel.grid = element_blank(),panel.background = element_blank())
}
# highlight_mask <- any(grepl("^Cpx",mcols(x)$manual_genesets))
# highlight_mask <- any(grepl("^CpxR",mcols(x)$regdb))
highlight_mask <- !any(grepl(".",mcols(x)$regdb))
p0 <- ggmap(x,~`log2_fc_avg.C005/C000`,~`log2_fc_avg.T005/C000`,highlight_mask) + ggtitle("0' >> 5'")
p1 <- ggmap(x,~`log2_fc_avg.C030/C005`,~`log2_fc_avg.T030/T005`,highlight_mask) + ggtitle("5' >> 30'")
p2 <- ggmap(x,~`log2_fc_avg.C120/C030`,~`log2_fc_avg.T120/T030`,highlight_mask) + ggtitle("30' >> 120'")
p0+p1+p2

plotly::ggplotly(p0,tooltip="text") %>% plotly::config(displayModeBar = TRUE)
plotly::ggplotly(p1,tooltip="text") %>% plotly::config(displayModeBar = TRUE)
plotly::ggplotly(p2,tooltip="text") %>% plotly::config(displayModeBar = TRUE)
```


# Show expression evolution accross time for selected genes

```{r}
gg_gene_timecourse <- function(X) {
  Z <- reshape2::melt(assay(X,"rpkm"))
  Z <- cbind(Z,colData(X)[Z$Var2,])
  Z <- cbind(Z,mcols(X)[Z$Var1,c("gene","locus_tag")])
  Z$gene <- factor(Z$gene,mcols(X)$gene)
  ggplot(Z) + facet_wrap(~gene,scales="free_y") +
    geom_line(aes(x=time,y=value,group=sample,color=cond)) + 
    geom_point(aes(x=time,y=value,color=cond)) + ylab("normalized gene expression (rpkm)") + 
    theme_classic()
}

gg_gene_timecourse(x[match(c("cpxP","ldtD","rpoE","lptA","eptA","rprA"),mcols(x)$gene)])
gg_gene_timecourse(x[c("cpxP","ldtD","rpoE","lptA","eptA","rprA")])
gg_gene_timecourse(x[match(c("degP","dsbA","ppiA","sulA"),mcols(x)$gene)])
gg_gene_timecourse(x[match(c("b3702"),mcols(x)$locus_tag)])
```


```{r,fig.width=6,fig.height=4}
gg_gene_timecourse(x[any(grepl("^Cpx stress",mcols(x)$manual_genesets))]) + ggtitle("Cpx stress")
gg_gene_timecourse(x[any(grepl("Riboflavin metabolism",mcols(x)$kegg_genesets))])
```


```{r,fig.width=3,fig.height=2.5}
gg_gene_timecourse(x[any(grepl("^Lipopolysaccharide transport",mcols(x)$manual_genesets))]) + ggtitle("^Lipopolysaccharide transport")
```




```{r fig.height=10, fig.width=5}
# To display heatmap of absolute normalize
gg_gene_timecourse_heatmap <- function(X) {
  Z <- assay(X,"rpkm")
  Z <- reshape2::melt(Z)
  Z <- cbind(Z,colData(X)[Z$Var2,])
  Z <- cbind(Z,mcols(X)[Z$Var1,c("gene","locus_tag")])
  Z$gene <- factor(Z$gene,mcols(X)$gene)
  ggplot(Z) + 
    geom_tile(aes(x=lib,y=gene,fill=value)) + 
    labs(fill="rpkm") +
    theme_classic() +
    theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
}

# To display heatmap of fold-changes
gg_gene_timecourse_heatmap_fc <- function(X,ref=c("C0","all"),lim=2) {
  ref <- match.arg(ref)
  Z <- assay(X,"rpkm")
  Z <- log2(Z + 1)
  if (ref=="all") {
    Z <- Z - rowMeans(Z)  
  } else {
    Z <- Z - rowMeans(Z[,paste0(X$cond,X$time)=="C0"])  
  }
  Z <- reshape2::melt(Z)
  Z <- cbind(Z,colData(X)[Z$Var2,])
  Z <- cbind(Z,mcols(X)[Z$Var1,c("gene","locus_tag")])
  Z$gene <- factor(Z$gene,mcols(X)$gene)
  
  Z$value[Z$value < -lim] <- -lim
  Z$value[Z$value > +lim] <- +lim
  
  ggplot(Z) + 
    geom_tile(aes(x=lib,y=gene,fill=value)) + 
    scale_fill_distiller(palette = "RdBu") +
    theme_classic() +
    theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
}
```


```{r}
X <- x[any(grepl("Riboflavin metabolism",mcols(x)$kegg_genesets))]
gg_gene_timecourse(X)
gg_gene_timecourse_heatmap_fc(X,ref="all") + ggtitle("Riboflavin metabolism")
gg_gene_timecourse_heatmap_fc(X,ref="C0") + ggtitle("Riboflavin metabolism")
```


```{r fig.height=10, fig.width=5}
X <- x[match(c("cpxP","ldtD","rpoE","lptA","eptA","rprA"),mcols(x)$gene)]

X <- x[any(mcols(x)$regdb %in% "SigE")]
X <- X[hclust(dist(t(scale(t(log2(assay(X,"rpkm")+1))))),method="ward.D2")$order]

gg_gene_timecourse(X)
gg_gene_timecourse_heatmap(X)
gg_gene_timecourse_heatmap_fc(X,ref="all")
gg_gene_timecourse_heatmap_fc(X,ref="C0")
```


# Regulon DB

```{r}
source("lib/gsea/lib_gsea.R")
gsea_hcol <- function(GSEA) {
  m <- -log10(GSEA$enrichment_pval)
  rownames(m) <- seq(nrow(m))
  m[m>7] <- 7
  U <- grepl("^up\\.",colnames(m))
  D <- grepl("^dn\\.",colnames(m))
  colnames(m) <- sub("^(up|dn)\\.","",colnames(m))
  gg_matrix_hcol2(m[,U],m[,D],GSEA$overlap_size[,U],GSEA$overlap_size[,D]) + 
    theme(legend.position = "none") + scale_fill_manual(values=c("#990099","yellow")) + 
    scale_y_continuous(breaks = seq(nrow(m)),labels = sprintf("%s (n=%d)",GSEA$geneset_name,GSEA$geneset_size))
}
mcols(x)$treatment_change <- cbind(
  "0' >> 5'" = (mcols(x)$log2_fc_avg[,"T005/C000"] - mcols(x)$log2_fc_avg[,"C005/C000"]),
  "5' >> 30'" = (mcols(x)$log2_fc_avg[,"T030/T005"] - mcols(x)$log2_fc_avg[,"C030/C005"]),
  "30' >> 120'" = (mcols(x)$log2_fc_avg[,"T120/T030"] - mcols(x)$log2_fc_avg[,"C120/C030"]),
  "5' >> 120'" = (mcols(x)$log2_fc_avg[,"T120/T005"] - mcols(x)$log2_fc_avg[,"C120/C005"])
)
M <- as.matrix(data.frame(check.names = FALSE,
  up = mcols(x)$treatment_change >= +log2(1.5),
  dn = mcols(x)$treatment_change <= -log2(1.5)
))
```


# Analysis of enrichment of regdb genesets
```{r,fig.height=6,fig.width=3}
gsea <- multi_gsea(M,setNames(mcols(x)$regdb,rownames(x)))
GSEA <- gsea[rowSums(gsea$enrichment_qval<=5e-2)>0,]
#GSEA <- subset(gsea,geneset_name %in% c("PhoB","PhoP","RcsB","RcsAB","GadE-RcsB","RcsB-BglJ"))
gsea_hcol(GSEA)
```

# Analysis of enrichment of KEGG genesets
```{r}
gsea <- multi_gsea(M,setNames(mcols(x)$kegg_genesets,rownames(x)))
GSEA <- gsea[rowSums(gsea$enrichment_qval<=5e-2)>0,]
gsea_hcol(GSEA)
```

# Analysis of enrichment of GOTERMS genesets
```{r,fig.height=30,fig.width=5}
gsea <- multi_gsea(M,setNames(mcols(x)$goall_term,rownames(x)))
GSEA <- gsea[rowSums(gsea$enrichment_qval<=5e-2)>0,]
gsea_hcol(GSEA)
```


